---
title: "2 | Understand Your Data"
output:
  pdf_document: default
  html_document: default
---
```{r echo=FALSE}
# If we don't want a chunk to be visible in the output, just insert 'echo = false'
# For example, we don't want any output of this chunk in our html or pdf document, but this chunk is necessary to run first
# because it initializes the working enviroment! (read all-data.csv)

# In an Rmd file, Markdown is outside the chunks and not R code.


# Functions to read in the CSV table that contains all the raw data.
# Before running these functions, make sure the file "all-data.csv" is
# in the local directory.
# Also, within the R environment, change the working directory to the directory
# that contains the data file using the toolbar menu:
# File -> Change dir
#

# Read the data from the csv file.
processors <- read.csv("all-data.csv")


################################################################
#
# This function returns the data from the desired column.
# Example:  clock<-get_column("Fp2000","Processor.Clock..MHz.")

get_column <- function(x,y) {

# x = string with the name of the desired benchmark
# y = desired column
#
# Find the indices of all rows that have an entry for the  
# indicated benchmark
benchmark <- paste(paste("Spec",x,sep=""),"..average.base.",
	sep="")
ix <- !is.na(processors[,benchmark])
return(processors[ix,y])
}
################################################################




################################################################
# This function extracts the interesting data columns for the given benchmark
# program and returns a dataframe with these columns.

extract_data <- function(benchmark) {

temp <- paste(paste("Spec",benchmark,sep=""),"..average.base.", sep="")

# perf = the performance reported in the database
perf <- get_column(benchmark,temp)

#nperf = performance normalized to the overall range
max_perf <- max(perf)
min_perf <- min(perf)
range <- max_perf - min_perf
nperf <- 100 * (perf - min_perf) / range

clock <- get_column(benchmark,"Processor.Clock..MHz.")
threads <- get_column(benchmark,"Threads.core")
cores <- get_column(benchmark,"Cores")
TDP <- get_column(benchmark,"TDP")
transistors <- get_column(benchmark,"Transistors..millions.")
dieSize <- get_column(benchmark,"Die.size..mm.2.")
voltage <- get_column(benchmark,"Voltage..low.")
featureSize <- get_column(benchmark,"Feature.Size..microns.")
channel <- get_column(benchmark,"Channel.length..microns.")
FO4delay <- get_column(benchmark,"FO4.Delay..ps.")
L1icache <- get_column(benchmark,"L1..instruction...on.chip.")
L1dcache <- get_column(benchmark,"L1..data...on.chip.")
L2cache <- get_column(benchmark,"L2..on.chip.")
L3cache <- get_column(benchmark,"L3..on.chip.")

return(data.frame(nperf,perf,clock,threads,cores,TDP,transistors,dieSize,voltage,featureSize,channel,FO4delay,L1icache,L1dcache,L2cache,L3cache))

}
################################################################


# Extract a new data frame for each of the benchmark programs available in the data set.

int92.dat <- extract_data("Int1992")
fp92.dat <- extract_data("Fp1992")
int95.dat <- extract_data("Int1995")
fp95.dat <- extract_data("Fp1995")
int00.dat <- extract_data("Int2000")
fp00.dat <- extract_data("Fp2000")
int06.dat <- extract_data("Int2006")
fp06.dat <- extract_data("Fp2006")

```


**G**OOD data is the basis of any sort of regression model, because we use
this data to actually construct the model. If the data is flawed, the
model will be flawed. It is the old maxim of _garbage in, garbage out._
Thus, the first step in regression modeling is to ensure that your data is
reliable. There is no universal approach to verifying the quality of your
data, unfortunately. If you collect it yourself, you at least have the advan-
tage of knowing its provenance. If you obtain your data from somewhere
else, though, you depend on the source to ensure data quality. Your job
then becomes verifying your source’s reliability and correctness as much
as possible.

## 2.1 || Missing Values
Any large collection of data is probably incomplete. That is, it is likely
that there will be cells without values in your data table. These missing
values may be the result of an error, such as the experimenter simply for-
getting to fill in a particular entry. They also could be missing because that
particular system configuration did not have that parameter available. For
example, not every processor tested in our example data had an L2 cache.
Fortunately, R is designed to gracefully handle missing values. R uses the
notation _NA_ to indicate that the corresponding value is not available.  
&nbsp;&nbsp;&nbsp;&nbsp;
Most of the functions in R have been written to appropriately ignore _NA_
values and still compute the desired result. Sometimes, however, you must
explicitly tell the function to ignore the NA values. For example, calling
the ```mean()``` function with an input vector that contains NA values causes it
to return _NA_ as the result. To compute the mean of the input vector while
ignoring the _NA_ values, you must explicitly tell the function to remove the
_NA_ values using ```mean(x, na.rm=TRUE)``` .

### 2.2 || Sanity Checking and Data Cleaning
Regardless of where you obtain your data, it is important to do some _sanity
checks_ to ensure that nothing is drastically flawed. For instance, you can
check the minimum and maximum values of key input parameters (i.e.,
columns) of your data to see if anything looks obviously wrong. One of
the exercises in Chapter 8 encourages you explore other approaches for
verifying your data. R also provides good plotting functions to quickly
obtain a visual indication of some of the key relationships in your data set.
We will see some examples of these functions in Section 3.1.  
&nbsp;&nbsp;&nbsp;&nbsp;
If you discover obvious errors or flaws in your data, you may have to
eliminate portions of that data. For instance, you may find that the perfor-
mance reported for a few system configurations is hundreds of times larger
than that of all of the other systems tested. Although it is possible that this
data is correct, it seems more likely that whoever recorded the data simply
made a transcription error. You may decide that you should delete those
results from your data. It is important, though, not to throw out data that
looks strange without good justification. Sometimes the most interesting
conclusions come from data that on first glance appeared flawed, but was
actually hiding an interesting and unsuspected phenomenon. This process
of checking your data and putting it into the proper format is often called
_data cleaning._  
&nbsp;&nbsp;&nbsp;&nbsp;
It also is always appropriate to use your knowledge of the system and
the relationships between the inputs and the output to inform your model
building. For instance, from our experience, we expect that the clock rate
will be a key parameter in any regression model of computer systems per-
formance that we construct. Consequently, we will want to make sure that
our models include the clock parameter. If the modeling methodology sug-
gests that the clock is not important in the model, then using the method-
ology is probably an error. We additionally may have deeper insights into
the physical system that suggest how we should proceed in developing a
model. We will see a specific example of applying our insights about the
effect of caches on system performance when we begin constructing more
complex models in Chapter 4.  
&nbsp;&nbsp;&nbsp;&nbsp;
These types of sanity checks help you feel more comfortable that your
data is valid. However, keep in mind that it is impossible to prove that
your data is flawless. As a result, you should always look at the results
of any regression modeling exercise with a healthy dose of skepticism and
think carefully about whether or not the results make sense. Trust your
intuition. If the results don’t feel right, there is quite possibly a problem
lurking somewhere in the data or in your analysis.

### 2.3 || The Example Data
I obtained the input data used for developing the regression models in the
subsequent chapters from the publicly available _CPU DB_ database [2].
This database contains design characteristics and measured performance
results for a large collection of commercial processors. The data was col-
lected over many years and is nicely organized using a common format and
a standardized set of parameters. The particular version of the database
used in this book contains information on 1,525 processors.  
&nbsp;&nbsp;&nbsp;&nbsp;
Many of the database’s parameters (columns) are useful in understand-
ing and comparing the performance of the various processors. Not all of
these parameters will be useful as predictors in the regression models, how-
ever. For instance, some of the parameters, such as the column labeled
_Instruction set width_, are not available for many of the processors. Oth-
ers, such as the _Processor family_, are common among several processors
and do not provide useful information for distinguishing among them. As
a result, we can eliminate these columns as possible predictors when we
develop the regression model.  
&nbsp;&nbsp;&nbsp;&nbsp;
On the other hand, based on our knowledge of processor design, we
know that the clock frequency has a large effect on performance. It also
seems likely that the parallelism-related parameters, specifically, the num-
ber of threads and cores, could have a significant effect on performance,
so we will keep these parameters available for possible inclusion in the
regression model.  
&nbsp;&nbsp;&nbsp;&nbsp;
Technology-related parameters are those that are directly determined by
the particular fabrication technology used to build the processor. The num-
ber of transistors and the die size are rough indicators of the size and com-
plexity of the processor’s logic. The feature size, channel length, and FO4
(fanout-of-four) delay are related to gate delays in the processor’s logic.
Because these parameters both have a direct effect on how much process-
ing can be done per clock cycle and effect the critical path delays, at least
some of these parameters could be important in a regression model that
describes performance.  
&nbsp;&nbsp;&nbsp;&nbsp;
Finally, the memory-related parameters recorded in the database are the
separate L1 instruction and data cache sizes, and the unified L2 and L3
cache sizes. Because memory delays are critical to a processor’s perfor-
mance, all of these memory-related parameters have the potential for being
important in the regression models.  
&nbsp;&nbsp;&nbsp;&nbsp;
The reported performance metric is the score obtained from the SPEC
CPU integer and floating-point benchmark programs from 1992, 1995,
2000, and 2006 [6–8]. This performance result will be the regression
model’s output. Note that performance results are not available for every
processor running every benchmark. Most of the processors have perfor-
mance results for only those benchmark sets that were current when the
processor was introduced into the market. Thus, although there are more
than 1,500 lines in the database representing more than 1,500 unique pro-
cessor configurations, a much smaller number of results are reported for
each individual benchmark.

### 2.4 || Data Frames

The _fundamental_ object used for storing tables of data in R is called a _data
frame._ We can think of a data frame as a way of organizing data into a
large table with a row for each system measured and a column for each
parameter. An interesting and useful feature of R is that all the columns
in a data frame do not need to be the same data type. Some columns may
consist of numerical data, for instance, while other columns contain textual
data. This feature is quite useful when manipulating large, heterogeneous
data files.  
&nbsp;&nbsp;&nbsp;&nbsp;
To access the CPU DB data, we first must read it into the R environ-
ment. R has built-in functions for reading data directly from files in the
csv (comma separated values) format and for organizing the data into data
frames. The specifics of this reading process can get a little messy, depend-
ing on how the data is organized in the file. We will defer the specifics of
reading the CPU DB file into R until Chapter 6. For now, we will use a
function called ```extract_data()``` , which was specifically written for reading
the CPU DB file.  
&nbsp;&nbsp;&nbsp;&nbsp;
To use this function, copy both the **all-data.csv** and **read-data.R** files
into a directory on your computer (you can download both of these files
from this book’s web site shown on p. ii). Then start the R environment
and set the local directory in R to be this directory using the _File -> Change
dir_ pull-down menu. Then use the _File -> Source R code_ pull-down menu
to read the **read-data.R** file into R. When the R code in this file completes,
you should have six new data frames in your R environment workspace:
```int92.dat , fp92.dat , int95.dat , fp95.dat , int00.dat , fp00.dat , int06.dat , ```
and ```fp06.dat``` .  
&nbsp;&nbsp;&nbsp;&nbsp;
The data frame ```int92.dat``` contains the data from the CPU DB database
for all of the processors for which performance results were available for
the SPEC Integer 1992 (Int1992) benchmark program. Similarly, fp92.dat
contains the data for the processors that executed the Floating-Point 1992
(Fp1992) benchmarks, and so on. I use the ```.dat``` suffix to show that the
corresponding variable name is a data frame.  
&nbsp;&nbsp;&nbsp;&nbsp;
Simply typing the name of the data frame will cause R to print the en-
tire table. For example, here are the first few lines printed after I type
```int92.dat``` , truncated to fit within the page:

             nperf       perf        clock   threads   cores  ...
-------     ------      --------   -------  -------   -----   ----
    1       9.662070    68.60000    100     1           1     ...
    2       7.996196    63.100000   125     1           1     ...
    3       16.363872   90.72647    166     1           1     ...
    4       13.720745   82.00000    175     1           1     ...
  ...
  
&nbsp;&nbsp;&nbsp;&nbsp;
The first row is the header, which shows the name of each column. Each
subsequent row contains the data corresponding to an individual processor.
The first column is the index number assigned to the processor whose data
is in that row. The next columns are the specific values recorded for that
parameter for each processor. The function ```head(int92.dat)``` prints out just
the header and the first few rows of the corresponding data frame. It gives
you a quick glance at the data frame when you interact with your data.  
&nbsp;&nbsp;&nbsp;&nbsp;
Table 2.1 shows the complete list of column names available in these
data frames. Note that the column names are listed vertically in this table,
simply to make them fit on the page.

Table 2.1: The names and definitions of the columns in the data frames  
&nbsp;&nbsp;&nbsp;&nbsp;
    containing the data from CPU DB.
    
Column Number     Column name       Definition
-------------     ------------      -------------------------
    1               (blank)         Processor index number
    2              nperf            Normalized performance
    3               perf            SPEC performance
    4               clock           Clock frequency (MHz)
    5               threads         Number of hardware threads available
    6               cores           Number of hardware cores available
    7               TDP             Thermal design power
    8             transistors       Number of transistors on the chip(M)
    9             dieSize           The size fo the chip
    10            voltage           Nominal operating voltage
    11            featureSize       Fabrication feature size
    12            channel           Fabrication channel size
    13            FO4delay          Fan-out-four delay
    14            L1icache          Level 1 instruction cache size
    15            L1dcache          Level 1 data cache size
    16            L2cache           Level 2 cache size
    17            L3cache           Level 3 cache size

### 2.5 || Accessing a Data Frame
We access the individual elements in a data frame using square brackets to
identify a specific cell. For instance, the following accesses the data in the
cell in row 15, column 12:

```{r}
int92.dat[15,12]
```
We can also access cells by name by putting quotes around the name:  
```{r}
int92.dat["71","perf"]
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  This expression returns the data in the row labeled 71 and the column
labeled perf . Note that this is not row 71, but rather the row that contains
the data for the processor whose name is 71 .  
  &nbsp;&nbsp;&nbsp;&nbsp;
  We can access an entire column by leaving the first parameter in the
square brackets empty. For instance, the following prints the value in every
row for the column labeled clock :  
```{r}
int92.dat[,"clock"]
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  Similarly, this expression prints the values in all of the columns for row
36:
```{r}
int92.dat[36,]
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  The functions nrow() and ncol() return the number of rows and columns,
respectively, in the data frame:  
```{r}
nrow(int92.dat)
ncol(int92.dat)
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  Because R functions can typically operate on a vector of any length, we
can use built-in functions to quickly compute some useful results. For ex-
ample, the following expressions compute the minimum, maximum, mean,
and standard deviation of the perf column in the int92.dat data frame:  
```{r}
min(int92.dat[,"perf"])
max(int92.dat[,"perf"])
mean(int92.dat[,"perf"])
sd(int92.dat[,"perf"])
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  This square-bracket notation can become cumbersome when you do a
substantial amount of interactive computation within the R environment.
R provides an alternative notation using the $ symbol to more easily access
a column. Repeating the previous example using this notation:  
```{r}
min(int92.dat$perf)
max(int92.dat$perf)
mean(int92.dat$perf)
sd(int92.dat$perf)
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  This notation says to use the data in the column named perf from the data
frame named int92.dat . We can make yet a further simplification using the
attach function. This function makes the corresponding data frame local to
the current workspace, thereby eliminating the need to use the potentially
awkward $ or square-bracket indexing notation. The following example
shows how this works:  
```{r}
attach(int92.dat)
min(perf)
max(perf)
mean(perf)
sd(perf)
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  To change to a different data frame within your local workspace, you
must first detach the current data frame:  
```{r}
detach(int92.dat)
attach(fp00.dat)
min(perf)
max(perf)
mean(perf)
sd(perf)
```
  &nbsp;&nbsp;&nbsp;&nbsp;
  Now that we have the necessary data available in the R environment,
and some understanding of how to access and manipulate this data, we are ready to generate our first regression model.